Detecting Manual Alterations in Biological Image Data Using Contrastive Learning and Pairwise Image Comparison

### 2025-04-03 paper/main.tex
Method (first version)

$$ \mathcal{D} = \{d_i \in [0, 256)^{l \times l \times 3}, i \in [0, N)\} $$

where $d_i$ is the RGB-decoded image, $l$ -- the length of an image side.

@@

Let $x$ and $y$ be two images ($x, y \in [0, 256)^{l \times l \times 3}$).
The model consists of two main parts:

@@

To train the encoder, let us define the loss function $\mathcal{L}$:

$$ \mathcal{L}(v_x, v_y, I(x, y)) \in [0, +\infty) $$

where $v_x$ and $v_y$ are the embeddings of images $x$ and $y$, respectively, 
and $I(x, y)$ is defined as follows:

$$ I(x, y) = \begin{cases} 
1, \text{if $x$ and $y$ are similar}, \\
0, \text{otherwise}
\end{cases} $$

@@

DRAFT

The problem of detecting reused biological and medical images is rooted in the complexity of
distinguishing between similar images while remaining invariant to various transformations.
The dataset consists of biological snapshots from publicly available sources, we will 
store them as an array $D$ with elements $d_i \in [0, 256)^{l \times l \times 3}$, where 
$l$ is the length of images sides. 
Let us define the \textit{solution} -- a method which meant to solve the stated problem.
Also we will use the termini \textit{model} -- the machine learning solution, which is 
proposed in our work.

Structure of the model is the same as in \texttt{Barlow Twins} \cite{zbontar2021barlow}.
It takes a batch of images $X$, each image $x_i$ is being augmented in two ways, producing 
two modified pictures $y_i^A$ and $y_i^B$ for each original. 
The resulted two batches of distorted images $Y^A$ and $Y^B$ are then
fed to a deep network function $f_{\theta}$, and the produced embeddings $Z^A$ and $Z^B$ 
are used to evaluate loss function $\mathcal{L}$ with the following formula, proposed in 
\texttt{Barlow Twins} \cite{zbontar2021barlow}:

$$ \mathcal{L}(Z^A, Z^B) = \sum_i {(1 - \zeta_{ii})^2} + \lambda \sum_{i}\sum_{j \neq i}{\zeta_{ij}^2} $$

where $\lambda$ is the positive constant, and $\zeta_{ij}$ -- cross-correlation matrix, 
computed between the outputs of the two identical networks along the batch dimension:

$$ \zeta_{ij} = 
\frac{\sum_b{z_{b,i}^A z_{b,j}^B}}{\sqrt{\sum_b{(z_{b,i}^A)^2}}\sqrt{\sum_b{(z_{b,j}^B)^2}}} $$

For the further simplicity, we will refer to a pair of images with the same content 
before alterations as a \textit{similar} pair, otherwise it will be called \textit{dissimilar}. 
The idea behind this loss function is to minimize the distance between similar pairs embeddings 
and to make it bigger for dissimilar.

The model is trained on the $70\%$ dataset $D$ data, then validated on another $10\%$ 
and tested on the remaining $20\%$ -- let $D_{test}$ be this part. For accuracy evaluation, 
we follow the widely used linear evaluation protocol (here comes the links from \cite{chen2020simclr}).

\section{Related Work}\label{sec:rw}

>>>>

$$ \mathcal{D} = \{d_i \in \mathcal{S}, i \in [0, N)\} $$

where $\mathcal{S}$ is the image space.

@@

Let $x$ and $y$ be two images ($x, y \in \mathcal{S}$).
The model consists of two main parts:

@@

%To train the encoder, let us define the loss function $\mathcal{L}$:

%$$ \mathcal{L}(v_x, v_y, I(x, y)) \in [0, +\infty) $$

%where $v_x$ and $v_y$ are the embeddings of images $x$ and $y$, respectively, 
%and $I(x, y)$ is defined as follows:

%$$ I(x, y) = \begin{cases} 
%1, \text{if $x$ and $y$ are similar}, \\
%0, \text{otherwise}
%\end{cases} $$

@@

The challenge of detecting reused biological and medical images lies in the difficulty of
distinguishing between visually similar images while maintaining invariance to various 
transformations. Our dataset consists of biological snapshots obtained from publicly available 
sources and is stored as an array $\mathcal{D}$ with elements $d_i \in [0, 256)^{l \times l 
\times 3}$, where $l$ denotes the length of the image sides. We define a \textit{solution} as 
any method intended to address the stated problem, and we refer to the \textit{model} as the 
machine learning solution proposed in our work.

The structure of our model is inspired by the \texttt{Barlow Twins} \cite{zbontar2021barlow}. 
It consists of three deep networks: an embedding function $f$, a projector function $p$, 
and a similarity function $s$ that returns a single number in the range $[0,1]$. 
The embedding network is implemented using a fine-tuned, pretrained \texttt{ResNet50}.

To train the projector, we begin with a batch of images $X$. Each image $x_i$ is 
augmented in two different ways to produce two modified versions, $y_i^A$ and $y_i^B$. 
These two batches of augmented images, $Y^A$ and $Y^B$, are then processed by the 
embedding function $f$ to yield embeddings $E^A$ and $E^B$. The embeddings are subsequently 
passed through the projector function $p$, resulting in the projected embeddings 
$Z^A$ and $Z^B$. These projected embeddings are used to compute the loss function 
$\mathcal{L}$ as proposed in \texttt{Barlow Twins} \cite{zbontar2021barlow}:

$$
\mathcal{L}(Z^A, Z^B) = \sum_i (1 - \mathcal{C}_{ii})^2 + \lambda \sum_{i} \sum_{j \neq i} \mathcal{C}_{ij}^2,
$$

where $\lambda$ is a positive constant, and $\mathcal{C}_{ij}$ is the cross-correlation matrix between the outputs of the two networks along the batch dimension, defined as:

$$
\mathcal{C}_{ij} = \frac{\sum_b z_{b,i}^A z_{b,j}^B}{\sqrt{\sum_b (z_{b,i}^A)^2} \sqrt{\sum_b (z_{b,j}^B)^2}}.
$$

For simplicity, we refer to a pair of images with the same content before augmentation as a 
\textit{similar} pair, and a pair with different content as a \textit{dissimilar} pair. The 
objective of the loss function is to minimize the distance between embeddings of similar pairs 
while maximizing the distance for dissimilar pairs.

For accuracy evaluation, we train the similarity network function $s$ while keeping the 
model's weights frozen. Given a batch of images $X$, we generate two batches, 
$Y^A$ and $Y^B$, where $Y^A$ comprises images from $X$ with random modifications, 
and $Y^B$ is a shuffled version of $X$ with different modifications applied. 
Consequently, for each image $x_i$, there is a corresponding modified image $y_i^A$ 
and a randomly transformed image $y_i^B$, with a probability of $0.3$ that they form a 
\textit{similar} pair. These batches are processed sequentially through the embedding 
function $f$ and the projector $p$, and the resulting embeddings $Z^A$ and $Z^B$ are then 
fed into the similarity function $s$. The function $s$ produces a vector $P$ with values 
in the range $[0,1]$, where each element $P_i$ represents the estimated likelihood that the 
pair $(y_i^A, y_i^B)$ is similar. The similarity network is trained using 
Binary Cross-Entropy (BCE) loss.

The model is trained on $70\%$ of the dataset $\mathcal{D}$, 
validated on an additional $10\%$, and tested on the remaining $20\%$.
